European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-23 manuscript/manuscript.qmd
add draft for 3rd and 4th para

As an international organization formally governed by a system of weighted voting and technocratic procedures, IMF is viewed as an “accountant” [@lutz2019], primarily concerned with the internal consistency of fiscal projections to guarantee timely repayment of its financial assistance. However, empirical studies suggested that countries with low international reserves, high current account deficits, and low GDP growth are more likely to receive IMF funding assistance [@barro2005; @knight1997], despite such indicators often reflecting a limited capacity to repay. Generally, all studies on IMF lending outcomes agree that major power impose their influence in the process. For example, studies have shown that countries strategically important to the United States are more likely to receive IMF loans, or to receive larger amounts. Specifically, this applies to countries that align with the U.S. in United Nations General Assembly and United Nations Security Council voting [@lipscy2018; @thacker1999; @dreher2009], are in geopolitical vital regions \[\], or hold significant debts to American commercial banks [@oatley2004]. This is not surprising, as the United States, as the most powerful state, has long projected its influence over international financial institutions—much like it has done with the World Bank [@wade2002; @woods2003; @clark2020]. The question this study brings about is, how this reflection of power is distributed within a given international organization, especially one that is widely perceived to be dominated by a few major powers.

>>>>

Quantitative studies also provide evidence of great power influence over IMF lending decisions. Such studies often use a variety of proxies to measure the recipient country's ties to the major powers, such as UN voting alignment [@thacker1999; @dreher2009], bilateral trade intensity [@barro2005], and financial exposure [@oatley2004]. These proxies aim to capture the strategic incentives of the great powers to support IMF lending to countries that align with its geopolitical interests. Such ties have been proven to be associated with more favourable IMF lending outcomes, such as larger loan amounts, higher participation rates, as well as less and more lenient conditionality [@lipscy2018; @barro2005; @dreher2009].

Yet, much of the quantitative literature has remained US-centric, often using only American alignment as the independent variable of interest [@thacker1999; @oatley2004; @dreher2009]. Others pool European indicators with US ones, often under the umbrella of “Western” influence [@lipscy2018, @barro2005].
